Enhancing Forecast Reconciliation: A Study of Alternative Covariance Estimators

### 2025-10-20 thesis/paper_draft.qmd
refine abstract & intro from George feedback

A collection of time series connected via a set of linear constraints is known as hierarchical time series. Forecasting these series without respecting the hierarchical nature of the data can lead to incoherent forecasts across aggregation levels and, in practice, reduced accuracy. Forecast reconciliation corrects this by adjusting base forecasts to satisfy such constraints. Among modern reconciliation methods, Minimum Trace (MinT) is widely used, however, it requires a good estimate of the forecast error covariance matrix. The current practice is to use the linear shrinkage toward a diagonal target to estimate 1-step-ahead covariance and proportionally scale it to approximate the h-step-ahead covariance matrix. This leaves a question of whether this method is appropriate for all real-world applications. We study alternative estimators that address the shortcomings of the current practice, including the NOVELIST estimator (shrinkage toward a soft-thresholded target), the PC-adjusted shrinkage (which utilises the latent factors structure), and horizon-specific estimators that relax proportional scaling. We evaluate MinT under these covariance estimates for both point and probabilistic reconciliation, and demonstrate their effectiveness and improvements over the shrinkage estimator in a complex, large-hierarchy dataset.

@@

In time series forecasting, aggregation occurs in a variety of settings. For example, Starbucks operates in many countries, and each country has multiple cities where they have outlets. The sales data is structured *hierarchically*: the top level is the total sales across all countries, which are the sum of sales in each country, which in turn are the sum of sales in each city, and finally the sum of sales from each outlet in the city. As a result, there are over 50,000 sales series across all *aggregation levels*, and decision makers need forecasts at each level to manage inventory and plan marketing strategies effectively. The hierarchy can be even more complex if we consider the sales of different kinds of drinks (e.g., coffees, teas, refreshers) at each aggregation level. Such hierarchical structures also arises in many other decision-making contexts, from supply chains [@Seaman2022-bb; @Angam2025-od] and energy planning [@Di-Modica2021-ad], to macroeconomics [@El-Gemayel2022-hx; @Li2019-vv] and tourism analysis [@Athanasopoulos2009-lp]. Stakeholders in these settings need forecasts at several aggregation levels to allocate resources and manage risk.

In practice, when forecasts are produced for all series (often called *base forecasts*), they typically violate the aggregation constraints observed in the data (e.g. the sum of all countries' sales does not equal the total sales). Such forecasts are called *incoherent*. Incoherence undermines downstream decisions that require internal consistency and can degrade forecasting performance. To tackle this problem, forecast reconciliation was introduced. Forecast reconciliation, a post-processing step, utilises the information from the hierarchical structure and data to adjust the initially produced base forecasts, so that the resulting forecasts are *coherent* (i.e. respecting the aggregation constraints). It first introduced by @Hyndman2011-jv, and later developed by @Van_Erven2015-nx, @Hyndman2016-ic, @Ben-Taieb2019-yx, @Wickramasuriya2019-xq, @Wickramasuriya2020-uk, and others to enhance point forecast performance. @Athanasopoulos2024-as provided a comprehensive review of the literature on forecast reconciliation. Among the modern methods, the Min Trace (MinT) approach developed by @Wickramasuriya2019-xq is is widely used due to its strong theoretical properties for minimising total reconciled error variance, computational efficiency, and robust empirical performance. MinT is implemented in popular R and Python software ecosystems [@O-Hara-Wild2024-we; @Nixtla].

A central difficulty for MinT is estimating the covariance matrix of base-forecast errors, particularly beyond one-step horizons. This is a high-dimensional estimation problem in which the number of series often exceeds the effective time dimension. A common practice, following @Wickramasuriya2019-xq, is to estimate the one-step covariance using linear shrinkage toward a diagonal target [@Schafer2005-yw], and to obtain multi-step covariances by proportional scaling of the one-step estimate. While convenient and guaranteed to produce a positive-definite estimator, this practice has three important shortcomings. First, the shrinkage is uniform across off-diagonals, applying a single penalty that may over-shrink genuine dependence and under-shrink noise. Second, many hierarchical data sets exhibit strong latent low-rank structure that is not explicitly exploited. Third, the proportionality relationship between the h-step-ahead and 1-step-ahead forecast error covariance matrices might not hold when error dynamics change with horizon.

@@

This paper focuses on covariance estimation for MinT. We examine the shortcomings of the standard shrinkage plus scaling practice and introduce alternative estimators that address these issues individually and in combination. Specifically, we study NOVELIST as a more adaptive target-based shrinkage; principal-component-adjusted (PC-adjusted) estimators that exploit latent factor structure; and horizon-specific estimators that relax proportional scaling, including direct estimation from multi-step residuals. We evaluate MinT under these alternatives for both point and probabilistic reconciliation. In a large, complex real-world hierarchy, we find: first, NOVELIST improves probabilistic performance relative to linear shrinkage, including higher empirical coverage; second, PC-adjusted estimators consistently outperform other estimators when common components are strong, for both point and probabilistic metrics; third, proportional scaling of the one-step covariance remains a competitive baseline, indicating that the assumption is not universally invalid.

The remainder of the paper is organised as follows. @sec-pre-theory sets out the framework for hierarchical time series, forecast reconciliation, and MinT, and motivates the need for improved covariance estimation. @sec-cov-est presents the covariance estimators considered, outlining their strengths and limitations in the reconciliation context. @sec-simulation details the simulation design and examines the performance of NOVELIST compared to shrinkage. @sec-empirical dives into a real-world application with all proposed methods and highlights behaviours not observed in the simplified simulations.

>>>>

A collection of time series connected via a set of linear constraints is known as hierarchical time series. Forecasting these series without respecting the hierarchical nature of the data can lead to incoherent forecasts across aggregation levels and, in practice, reduced accuracy. Forecast reconciliation corrects this by adjusting base forecasts to satisfy such constraints. Among modern reconciliation methods, Minimum Trace (MinT) is widely used, however, it requires a good estimate of the forecast error covariance matrix. The current practice is to use the linear shrinkage towards a diagonal target. Furthermore, the covariance estimate is based on 1-step-ahead residuals, then proportionally scale it to approximate the h-step-ahead covariance matrix. This leaves a question of whether this method is appropriate for all real-world applications. We study the shortcomings of current practice and propose alternative covariance estimators, including the NOVELIST estimator (shrinkage towards a soft-thresholded target), PC-adjusted shrinkage (which utilises latent factor structures), and horizon-specific estimators that relax proportional scaling. We evaluate MinT using these covariance estimates for both point and probabilistic reconciliation, and demonstrate their effectiveness and improvements over the shrinkage estimator in a complex, large-hierarchy dataset.

@@

In time series forecasting, aggregation occurs in a variety of settings. For example, Starbucks Corporation operates in many countries, and each country has multiple cities where they have outlets. The sales data is structured *hierarchically*: the top level is the company's total sales, which disaggregates into sales by country, then sales by city within each country, and finally down to sales by individual outlet within each city. As a result, there are over 50,000 sales series across all *aggregation levels*, and decision makers need forecasts at each level to manage inventory and plan marketing strategies effectively. The hierarchy can be even more complex if we consider the sales of different categories of products (e.g., beverage, food, etc.) and the sales of product within each category (e.g., latte, cappuccino, etc.) at each aggregation level. In this case, the structure is called a *grouped structure*, where the aggregation paths are not unique. Such structures also arises in many other decision-making contexts, from supply chains [@Seaman2022-bb; @Angam2025-od] and energy planning [@Di-Modica2021-ad], to macroeconomics [@El-Gemayel2022-hx; @Li2019-vv] and tourism analysis [@Athanasopoulos2009-lp]. Stakeholders in these settings need forecasts at several aggregation levels to allocate resources and manage risk.

In practice, when forecasts are produced for all series (often called *base forecasts*), they typically violate the aggregation constraints observed in the data (e.g., the sum of all countries' sales forecasts does not equal the total sales forecast). Such forecasts are called *incoherent*. Incoherence undermines downstream decisions that require internal consistency and can degrade forecasting performance. To tackle this problem, forecast reconciliation was introduced. Forecast reconciliation, a post-processing step, utilises the information from the hierarchical structure and data to adjust the initially produced base forecasts, so that the resulting *reconciled forecasts* are *coherent* (i.e., respecting the aggregation constraints). It was first introduced by @Hyndman2011-jv, and later developed by @Van_Erven2015-nx, @Hyndman2016-ic, @Ben-Taieb2019-yx, @Wickramasuriya2019-xq, @Wickramasuriya2020-uk, and others that focus point forecast reconciliation. Recognising the important of probabilistic forecasting, probabilistic forecast reconciliation was studied by @Shang2017-ux, @Jeon2019-wm, and @Ben-Taieb2021-bn, and later formally defined by @Panagiotelis2023-fs. @Athanasopoulos2024-as provide a comprehensive review of the literature on forecast reconciliation. Among the modern methods, the Min Trace (MinT) approach developed by @Wickramasuriya2019-xq is is widely used due to its strong theoretical properties for minimising total reconciled forecast error variance, computational efficiency, and robust empirical performance. MinT is later extended to probabilistic reconciliation by @Wickramasuriya2024-mb, showing that it also minimises the negative log score of the reconciled distribution under Gaussian assumptions. MinT is implemented in popular R and Python software ecosystems [@O-Hara-Wild2024-we; @Nixtla].

A central difficulty for MinT is estimating the covariance matrix of base-forecast errors, particularly beyond one-step forecast horizons. This is a high-dimensional estimation problem in which the number of series often exceeds the time dimension. A common practice, following @Wickramasuriya2019-xq, is to estimate the 1-step-ahead covariance from the residuals, using linear shrinkage toward a diagonal target [@Schafer2005-yw]. Then, proportionally scale this estimate to approximate the multi-step-ahead covariance matrix. While convenient and guaranteed to produce a positive-definite estimate, this practice has three important shortcomings. First, the shrinkage is uniform across off-diagonals, applying a single penalty that may over-shrink genuine dependence and under-shrink noise. Second, many hierarchical data sets exhibit strong latent low-rank structures that can be explicitly exploited. Third, the proportionality relationship between the h-step-ahead and 1-step-ahead forecast error covariance matrices might not hold when error dynamics change with horizon.

@@

This paper focuses on covariance estimation for MinT. We examine the shortcomings of the standard shrinkage plus scaling practice and introduce alternative estimators that address these issues individually and in combination. Specifically, we study NOVELIST as a more adaptive target-based shrinkage; principal-component-adjusted (PC-adjusted) estimators that exploit latent factor structures; and horizon-specific estimators that relax proportional scaling, including direct estimation from multi-step residuals. We evaluate MinT under these alternatives for both point and probabilistic reconciliation. In a large, complex real-world hierarchy, our findings reveal three main insights. First, NOVELIST improves probabilistic performance relative to linear shrinkage, including higher empirical coverage. Second, PC-adjusted estimators consistently outperform other estimators when common components are strong, for both point and probabilistic metrics. Third, proportional scaling of the one-step covariance remains a competitive baseline, indicating that the assumption may not be universally invalid.

The remainder of the paper is organised as follows. @sec-pre-theory sets out the framework for hierarchical time series, forecast reconciliation, and MinT, and motivates the need for improved covariance estimation. @sec-cov-est presents the covariance estimators considered, outlining their strengths and limitations in the reconciliation context. @sec-scores defines the evaluation metrics for point and probabilistic reconciliation. @sec-simulation details the simulation design and examines the performance of NOVELIST compared to shrinkage. @sec-empirical dives into a real-world application with all proposed methods and highlights behaviours not observed in the simplified simulations.
